---
title: "STA457 Project"
output: pdf_document
---

```{r, echo=TRUE, results='hide', message=F, warning=F}
library(tidyverse)
library(forecast)
library(astsa)
library(tseries)
```

## Read Data

```{r, echo=TRUE, results='hide', message=F, show_col_types = FALSE, warning=F}
data <- read_csv("data/cleaned_data.csv")

price_ts <- ts(data$PriceAvg, start=c(1994, 10), frequency=12)
temp_ts <- ts(data$PrcpAvg, start=c(1994, 10), frequency=12)
rain_ts <- ts(data$TAvg, start=c(1994, 10), frequency=12)
currency_ts <- ts(data$Dom_USD_Period_Average, start=c(1994, 10), frequency=12)
```

## Plot

```{r}
plot.ts(price_ts)
plot.ts(temp_ts)
plot.ts(rain_ts)
plot.ts(currency_ts)
```

## Checking Assumptions

```{r}
acf2(price_ts)
acf2(temp_ts)
acf2(rain_ts)
acf2(currency_ts)

adf.test(price_ts)
adf.test(temp_ts)
adf.test(rain_ts)
adf.test(currency_ts)
```

From the data, we can see that none of these time series are stationary. Specifically, the temperature and precipitation time series have a p-value of 0.01, so we fail to reject the null hypothesis. Therefore, the time series are not stationary.

We'll try differencing the temperature and rain times series in order to create an ARIMA model off of that.

```{r}
log_price <- log(price_ts)
diff_price <- diff(log_price)
acf2(diff_price)
```

After differencing, we can see a significant spike in the ACF plot at lag 1 that later tails off, and in the PACF plot, we can see a significant spike at lag 1 that tails off. As a result, we will try fitting an ARIMA(1,1,0) model, an ARIMA(0,1,1) model, and an ARIMA(1,1,1) to see which one has a lower AIC value.

## Univariate ARIMA Models

```{r}
# ARIMA model
arima_model1 <- arima(log_price, order = c(1,1,0))
arima_model2 <- arima(log_price, order = c(0,1,1))
arima_model3 <- arima(log_price, order = c(1,1,1))
AIC(arima_model1, arima_model2, arima_model3)
```

# Checking Residuals

```{r}
checkresiduals(arima_model1) # ARIMA(1,1,0) model
checkresiduals(arima_model2) # ARIMA(0,1,1) model
checkresiduals(arima_model3) # ARIMA(1,1,1) model
```
