---
Author: "Ashka Shah"
Date: "2025-03-28"
---
# Regession with lagged variables

```{r}
library(tseries) 
library(ggplot2)
library(forecast)
```

Let's load the data and check whether all the series are stationary by looking at their ACFs If they aren't we will need to transform them.

```{r}

data_x <- read.table("data/cleaned_data.csv", sep=",", header=TRUE)

# Get the separate time series
temp <- data_x$TAvg
precip <- data_x$PrcpAvg
price <- data_x$PriceAvg
idx <- data_x$Dom_USD_Period_Average

#plot the ACFs

acf(temp, main="ACF of Temperature Series")
acf(precip, main="ACF of Precipitation Series")
acf(price,main="ACF of Futures Price Series")
acf(idx,  main="ACF of Cedi Exchange Rate Series")

```

All series are non-stationary, so let's apply differencing to make them stationary. The Temperature and Precipitation series seems to show annual seasonality so we can difference it accordingly.

```{r}
l_temp <-  diff(temp, lag = 12) # There seems to be some annual seasonality
l_precip <- diff(precip, lag = 12) # There seems to be some annual seasonality
l_price <- diff(log(price))
l_idx <- diff(log(idx))

#plot the ACFs
acf(l_temp, main="ACF of Diff Temperature Series")
acf(l_precip, main="ACF of Diff Precipitation Series")
acf(l_price, main="ACF of Diff Futures Price Series")
acf(l_idx,  main="ACF of Diff Cedi Exchange Rate Series")

```



All the series are now stationary. We can continue with finding Cross-correlations between the Price series and the other three series.

```{r}
ccf(l_price, l_temp, lag.max = 25, main = "Cross-Correlation between Price and Temperature")
ccf(l_price, l_precip, lag.max = 25, main = "Cross-Correlation between Price and Precipitation")
ccf(l_price, l_idx, lag.max = 25, main = "Cross-Correlation between Price and Currency Rates")
```

Here are some observations from the CCF graphs - 1) There seems to be a significant cross correlation between Price and Currency Rates at Lag 17. 2) There seems to be a significant cross correlation between Price and Precipitation at Lag 22. 3) There seems to be a significant cross correlation between Price and Temperature at Lag 24.

Based on the previous ACF graphs, Prices seems to have a co-relation with Lag 1 and 7 Price. So, we will also use that. We don't want to include all the lags as it increases the number of variables the model has to predict. As a result of adding lags to the data and the differencing for precipitation and temperature we will lose the initial 35 months of data.

```{r}
# Editing the length of temp and precip

l_temp <-  c(rep(NA, 11),l_temp) # There seems to be some annual seasonality
l_precip <- c(rep(NA, 11),l_precip) # There seems to be

# Lagging the data
lag_cr_17 <- c(rep(NA, 17), l_idx[1:(length(l_idx) - 17)])
lag_pr_1 <- c(rep(NA, 1), l_price[1:(length(l_price) - 1)])
lag_pr_7 <- c(rep(NA, 7), l_price[1:(length(l_price) - 7)])
lag_pcp_22 <- c(rep(NA, 22), l_precip[1:(length(l_precip) - 22)])
lag_tmp_24 <- c(rep(NA, 24), l_temp[1:(length(l_temp) - 24)])

# Let's drop the first 35 rows as lag 24 is our max lag and we already diff the temp/prcp series with lag 12. So, we will have data only from 1997 Aug.
edited_data <- cbind(l_price, l_temp, l_precip, l_idx, lag_pr_1, lag_pr_7, lag_cr_17, lag_tmp_24, lag_pcp_22)
```

Now, let's split this dataset into test and validation. We will use the last 4 data points for test set.

```{r}
test_set <- data.frame(tail(edited_data,4))
train_set <- head(edited_data, n = nrow(edited_data) - 4)
# Now let's adjust the train_set for lags
train_set <- data.frame(edited_data[36:nrow(train_set), ])
head(train_set)

model <- lm(l_price ~ lag_pr_1 + lag_pr_7 + lag_cr_17 + lag_pcp_22 + lag_tmp_24, data=train_set)

```

```{r}
summary(model)
AIC(model)

checkresiduals(model)
plot(model$residuals)

# Q-Q plot of residuals
qqnorm(residuals(model))
qqline(residuals(model))

Box.test(residuals(model),lag=10, type = "Ljung-Box")

```
Let's forecast the last 4 values.

```{r}  
forecasted <- predict(model, newdata = test_set[, -which(names(test_set) == "l_price")])
print(forecasted)
```
Before we plot the forecasted prices we need to adjust for the differencing and log transformations.

```{r}

for_prices  <- c(log(data_x$PriceAvg[length(data_x$PriceAvg)-3])) # Get the last 5th log price from the training set

for (i in 1:length(forecasted)) {
  for_prices <- c(for_prices, for_prices[length(for_prices)] + forecasted[i])
}
for_prices <- for_prices[-1] # Removing the first price
for_prices <- exp(for_prices)

rmse <- sqrt(mean((for_prices - data_x$PriceAvg[(length(data_x$PriceAvg)-3):length(data_x$PriceAvg)])^2))
print(paste("RMSE: ", rmse))

print("Forecasted Prices")
print(for_prices)
print("Actual Prices")
print(data_x$PriceAvg[(length(data_x$PriceAvg)-3):length(data_x$PriceAvg)])

# Plot Original vs Forecasted Values
for_prices <- c(rep(NA, length(data_x$PriceAvg) - length(for_prices)), for_prices)
ggplot() +
  geom_line(aes(x=as.Date(data_x$YearMonth),y=data_x$PriceAvg), color = "black") +# Line plot
  geom_line(aes(x=as.Date(data_x$YearMonth),y=for_prices), color = "red") +
  labs( x = "Date", y = "Cocoa Future Prices")
```

